---
title: "Assignment 4"
author: "Andreas"
date: "2024-04-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(data.table)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(cmdstanr)
library(posterior)
```



In this markdown

- Simulate task environment
- Create agent and simulate data
- Inspect simulated behaviour
- Fit Stan model to simulated data
  - Fit quality check
- Fit Stan model to empirical data
  - Fit quality check
  

### Simulate task environt

```{r}
n <- 5
l <- rep(list(0:1), n)
stim <- expand.grid(l)

stim$dangerous <- ifelse(stim$Var1 == 1 | stim$Var2 == 1, 1, 0) # try with or here
stim$nutricious <- ifelse(stim$Var4 == 1, 1, 0)
stim$category <- ifelse(stim$dangerous == 0 & stim$nutricious == 0, 1,
                        ifelse(stim$dangerous == 0 & stim$nutricious == 1, 2,
                               ifelse(stim$dangerous == 1 & stim$nutricious == 0, 3, 4)))


stim_full <- data.frame()

for (i in 1:3){ # try 6
  stim_full <- rbind(stim_full, stim[sample(nrow(stim)),])
}

stim_full$response = NA


```


### Create agent

```{r}
# GCM_f <- function(c, w1, w2, w3, w4, w5, data){
#   data$response <- NA
#   
#   ntrials <- nrow(data)
#   
#   categories <- c(1,2,3,4)
#   
#   observed <- c()
#   
#   pos <- array(NA, dim=c(4,5)) # position array to update. Dimension 4, because we have 4 categories
#   
#   
#   
#   for (t in 1:ntrials){
#     if (sum(c(1,2,3,4) %in% observed != 4)){ # We check if each category has been observed. We need 4 TRUEs
#       data$response[t] <- sample(categories, 1, replace = TRUE) # If we havent observed all cats, we guess
#     }
#     else{
#       cat_dist <- array(NA, dim=4) # Distance to each category, empty array
#       cat_sim <- array(NA, dim=4) # Similarity with each category, empty array
#       
#       for (cat in 1:4){
#         fwise_dist <- array(NA, dim=5)
#         
#         fwise_dist[1] <- data$Var1[r] - pos[cat,1]
#         fwise_dist[2] <- data$Var2[r] - pos[cat,2]
#         fwise_dist[3] <- data$Var3[r] - pos[cat,3]
#         fwise_dist[4] <- data$Var4[r] - pos[cat,4]
#         fwise_dist[5] <- data$Var5[r] - pos[cat,5]
#         
#         # Calculate overall distance to category
#         cat_dist[cat] <- sqrt(w1*fwise_dist[1]^2 + w2*fwise_dist[2]^2
#                               + w3*fwise_dist[3]^2 + w4*fwise_dist[4]^2 + w5*fwise_dist[5]^2)
#         
#         # Calculate similarity to category
#         cat_sim[cat] <- exp(-c*cat_dist[cat])
#         
#         # burde vi dele beslutningen op i to? dvs først regne sim ift danger og derefter regne sim ift nutrition? hmmmm
#         }
#       }
# 
#       
#       
#     }
#     # Update pos
#   }
# }
```



### Create agent take 2

```{r}
GCM_f <- function(c, weights, data, R){
  
  data$response <- NA
  
  ntrials <- nrow(data)
  
  #categories <- c("d0", "d1", "n0", "n1")
  
  observed <- c()
  
  # We need arrays for both P and K (for all 4 categories), so we can go back.
  # We must also copy the old position value into the current trial, if that category is not updated
  
  dang0_pos <- array(NA, dim=c(ntrials+1,5))
  dang1_pos <- array(NA, dim=c(ntrials+1,5))
  nutr0_pos <- array(NA, dim=c(ntrials+1,5))
  nutr1_pos <- array(NA, dim=c(ntrials+1,5))
  
  # Defining initial position as 0.5 on all features axes
  dang0_pos[1,] <- rep(0.5, 5)
  dang1_pos[1,] <- rep(0.5, 5)
  nutr0_pos[1,] <- rep(0.5, 5)
  nutr1_pos[1,] <- rep(0.5, 5)
  
  # Defining K arrays (gain)
  # We dont need initial values for these, just that the arrays are same length
  dang0_K <- array(NA, dim = ntrials+1)
  dang1_K <- array(NA, dim = ntrials+1)
  nutr0_K <- array(NA, dim = ntrials+1)
  nutr1_K <- array(NA, dim = ntrials+1) 
  
  # Defining P arrays
  dang0_P <- array(NA, dim = ntrials+1)
  dang1_P <- array(NA, dim = ntrials+1)
  nutr0_P <- array(NA, dim = ntrials+1)
  nutr1_P <- array(NA, dim = ntrials+1)
  
  # Initial values of P are set to 1
  dang0_P[1] <- 1
  dang1_P[1] <- 1
  nutr0_P[1] <- 1
  nutr1_P[1] <- 1
  
  # Define distance function
  distance <- function(vec1, vec2, weights){
    return(sqrt(sum(weights * abs(vec1 - vec2)^2)))
  }
  
  # Define similarity function
  similarity <- function(distance, c){
    return(exp(-c * distance))
  }
  
  for (t in 1:ntrials){
    
    # The current alien; list of the features it has
    alien <- array(c(data$Var1[t],
                     data$Var2[t],
                     data$Var3[t],
                     data$Var4[t],
                     data$Var5[t]))
    
    if ("n1" %in% observed & "n0" %in% observed){
       # Calculate distances for the nutrition axis
      nutr0_dist <- distance(alien, nutr0_pos[t,], weights)
      nutr1_dist <- distance(alien, nutr1_pos[t,], weights)
      
      nutr0_sim <- similarity(nutr0_dist, c)
      nutr1_sim <- similarity(nutr1_dist, c)
      
      resp_nutr <- rbinom(1,1, nutr1_sim / (nutr1_sim + nutr0_sim))
      
    }
    else{
      resp_nutr <- rbinom(1,1, 0.5)
    }
    
    if ("d1" %in% observed & "d0" %in% observed) {
      # Calculate distances for the danger axis
      dang0_dist <- distance(alien, dang0_pos[t,], weights)
      dang1_dist <- distance(alien, dang1_pos[t,], weights)
      
      # Calculate similarities
      dang0_sim <- similarity(dang0_dist, c)
      dang1_sim <- similarity(dang1_dist, c)
      
      # Now code the response. We decide danger and nutrition seperately
      resp_dang <- rbinom(1,1, dang1_sim / (dang1_sim + dang0_sim))
 
    }
    else{
      resp_dang <- rbinom(1,1,0.5)
    }
    
    data$response[t] <- 1 + resp_nutr + 2 * resp_dang

    # append category to observed list
    observed <- c(observed, ifelse(data$dangerous[t] == 1, "d1", "d0"))
    observed <- c(observed, ifelse(data$nutricious[t] == 1, "n1", "n0"))
    
    # Update positions using Kalman filter

    # Update each of the four positions individually, as they are in principle independent.
    # Their independence is not clear, when we are in binary space, but on a continuous space, there could
    # be a sweetspot that meant danger. You could also imagine that eyes on stalk means danger sometimes
    # and eyes on stalk in combination with a specific other feature means not danger. Then the position of
    # both dang1 and dang0 would be towards the 1 end of the eyes on stalk axis.
    
    # Shifting the index with 1, so we can always go backwards
    k = t + 1
    
    if (data$dangerous[t] == 1){
      #kalman update dang1_pos
      dang1_K[k] = dang1_P[k-1] / (dang1_P[k-1] + R) # gain
      dang1_pos[k,] = dang1_pos[k-1,] + dang1_K[k] * (alien - dang1_pos[k-1,]) # update position
      dang1_P[k] = (1 - dang1_K[k]) * dang1_P[k-1] # update P
      
      
      #repeat values for dang0
      dang0_K[k] = dang0_K[k-1]
      dang0_pos[k,] = dang0_pos[k-1,]
      dang0_P[k] = dang0_P[k-1]
      
    }
    else{
      #kalman update dang0_pos
      dang0_K[k] = dang0_P[k-1] / (dang0_P[k-1] + R) # gain
      dang0_pos[k,] = dang0_pos[k-1,] + dang0_K[k] * (alien - dang0_pos[k-1,]) # update position
      dang0_P[k] = (1 - dang0_K[k]) * dang0_P[k-1] # update P
      
      
      #repeat values for dang1
      dang1_K[k] = dang1_K[k-1]
      dang1_pos[k,] = dang1_pos[k-1,]
      dang1_P[k] = dang1_P[k-1]

    }
    
    
    if (data$nutricious[t] == 1){
      #kalman update nutr1_pos
      nutr1_K[k] = nutr1_P[k-1] / (nutr1_P[k-1] + R) # gain
      nutr1_pos[k,] = nutr1_pos[k-1,] + nutr1_K[k] * (alien - nutr1_pos[k-1,]) # update position
      nutr1_P[k] = (1 - nutr1_K[k]) * nutr1_P[k-1] # update P
      
      
      #repeat values for nutr0
      nutr0_K[k] = nutr0_K[k-1]
      nutr0_pos[k,] = nutr0_pos[k-1,]
      nutr0_P[k] = nutr0_P[k-1]
    }
    else{
      #kalman update nutr0_pos
      nutr0_K[k] = nutr0_P[k-1] / (nutr0_P[k-1] + R) # gain
      nutr0_pos[k,] = nutr0_pos[k-1,] + nutr0_K[k] * (alien - nutr0_pos[k-1,]) # update position
      nutr0_P[k] = (1 - nutr0_K[k]) * nutr0_P[k-1] # update P
      
      
      #repeat values for nutr1
      nutr1_K[k] = nutr1_K[k-1]
      nutr1_pos[k,] = nutr1_pos[k-1,]
      nutr1_P[k] = nutr1_P[k-1]
    }
    
  }
  return(list(data = data,
              dang1_pos = dang1_pos,
              dang0_pos = dang0_pos,
              nutr1_pos = nutr1_pos,
              nutr0_pos = nutr0_pos))
}
```


Simulate data

```{r}
c = 3.5 # Higher c -> more deterministic. We need pretty high c, for agent to perform above chance
weights = array(rep(0.2, 5))
R = 10 # Slow learning
#R = 0.1 # Fast learning
sim <- GCM_f(c, weights, stim_full, R)

sim_d <- sim$data

dang1_d <- as.data.frame(sim$dang1_pos)
dang0_d <- as.data.frame(sim$dang0_pos)
nutr1_d <- as.data.frame(sim$nutr1_pos)
nutr0_d <- as.data.frame(sim$nutr0_pos)

```



Visualize positions over time

```{r}
p1 <- dang1_d %>% mutate(trial = seq(length(dang1_d$V1))) %>%
  mutate(V2 = V2 - 0.008) %>% 
  pivot_longer(cols = c(V1, V2, V3, V4, V5), values_to = "position", names_to = "feature") %>% 
  ggplot() +
  geom_line(aes(x = trial, y = position, color = feature), alpha = 0.5, size = 1.5) +
  ggtitle("Dang1")


p2 <- dang0_d %>% mutate(trial = seq(length(dang0_d$V1))) %>%
  mutate(V2 = V2 - 0.008) %>% 
  pivot_longer(cols = c(V1, V2, V3, V4, V5), values_to = "position", names_to = "feature") %>% 
  ggplot() +
  geom_line(aes(x = trial, y = position, color = feature), alpha = 0.5, size = 1.5)+
  ggtitle("Dang0")


p3 <- nutr1_d %>% mutate(trial = seq(length(nutr1_d$V1))) %>%
  pivot_longer(cols = c(V1, V2, V3, V4, V5), values_to = "position", names_to = "feature") %>% 
  ggplot() +
  geom_line(aes(x = trial, y = position, color = feature), alpha = 0.5, size = 1.5) +
  ggtitle("Nutr1")


p4 <- nutr0_d %>% mutate(trial = seq(length(nutr0_d$V1))) %>%
  pivot_longer(cols = c(V1, V2, V3, V4, V5), values_to = "position", names_to = "feature") %>% 
  ggplot() +
  geom_line(aes(x = trial, y = position, color = feature), alpha = 0.5, size = 1.5)+
  ggtitle("Nutr0")

(p1 + p2) / (p3 + p4)

```


```{r}
# write.csv(dang0_d, "dang0.csv")
# write.csv(dang1_d, "dang1.csv")
# write.csv(nutr0_d, "nutr0.csv")
# write.csv(nutr1_d, "nutr1.csv")
```


Visualize proportion of correct over time

```{r}
sim_d$correct <-ifelse(sim_d$response == sim_d$category, 1, 0)

sim_d$cummulativeRate <- cumsum(sim_d$correct) / seq_along(sim_d$correct)

sim_d %>% ggplot(aes(x = seq_along(cummulativeRate), y = cummulativeRate)) +
  geom_line() +
  ggtitle("Cummulative Rate") +
  xlab("Trial") +
  theme_minimal()

```


### Simulation with various combinations for R and C parameters


```{r}
Rs <- c(0.1, 0.5, 0.75, 1, 1.5, 2, 2.5, 3, 3.5, 5, 10)
Cs <- c(1, 1.5, 2, 2.5, 3, 3.5, 4)
weights = array(rep(0.2, 5))

d <- data.frame()

for (R in Rs){
  for (c in Cs){
    sim_temp <- GCM_f(c, weights, stim_full, R) # Simulate data
    temp_d <- sim_temp$data
    
    temp_d$correct <-ifelse(temp_d$response == temp_d$category, 1, 0)
    temp_d$cumulativeRate <- cumsum(temp_d$correct) / seq_along(temp_d$correct) # calc cummulative rate
    
    temp_d$R <- R
    temp_d$c <- c
    temp_d$Trial <- seq_along(temp_d$correct)
    
    d <- rbind(d, temp_d)
    
  }
}

```


Visualize performance by R and c

```{r}
d %>%
  ggplot(aes(x = Trial, y = cumulativeRate, group = c, color = c)) +
  geom_line() +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "red") +
  facet_wrap(~R) +
  ggtitle("CumulativeRate - facets by R") +
  theme_minimal()

d %>%
  ggplot(aes(x = Trial, y = cumulativeRate, group = R, color = R)) +
  geom_line() +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "red") +
  facet_wrap(~c) +
  ggtitle("CumulativeRate - facets by c") +
  theme_minimal()
```


### Fit model to simulated data


```{r}
# prepare data
w_prior_values = c(rep(0.2, 5))
ntrials = length(sim_d$Var1)
nfeats = 5
stimuli = array(NA, dim = c(ntrials, nfeats))
stimuli[,1] = sim_d$Var1
stimuli[,2] = sim_d$Var2
stimuli[,3] = sim_d$Var3
stimuli[,4] = sim_d$Var4
stimuli[,5] = sim_d$Var5

# prepare model
model_file = file.path("C:/Users/andre/OneDrive - Aarhus universitet/Cognitive Science/Kandidat/Advanced Cognitive Modeling/Assignments/Ass4/ass_4_GCMKalman.stan")

gcm_kalman_mod <- cmdstan_model(model_file)

data <- list(
  ntrials = ntrials,
  nfeats = nfeats,
  stimuli = stimuli,
  response = sim_d$response,
  dangerous = sim_d$dangerous,
  nutricious = sim_d$nutricious,
  w_prior_values = w_prior_values
)

# fit model

samples <- gcm_kalman_mod$sample(
  data = data,
  seed = 123,
  chains = 3,
  iter_warmup = 1000,
  iter_sampling = 3000,
  refresh = 0,
  max_treedepth = 10,
  adapt_delta = 0.80
)
```
inspect posteriors
```{r}
draws <- as_draws_df(samples$draws())
```


```{r}
draws %>% ggplot +
  geom_density(aes(x = logit_c), color = "red", fill = "red", alpha = 0.2) +
  geom_density(aes(x = logit_c_prior), color = "blue", fill = "blue", alpha = 0.2) +
  theme_minimal()

draws %>% ggplot +
  geom_density(aes(x = log_R), color = "red", fill = "red", alpha = 0.2) +
  geom_density(aes(x = log_R_prior), color = "blue", fill = "blue", alpha = 0.2) +
  theme_minimal()

draws %>% ggplot +
  geom_density(aes(x = c), color = "red", fill = "red", alpha = 0.2) +
  geom_density(aes(x = c_prior), color = "blue", fill = "blue", alpha = 0.2) +
  theme_minimal()

draws %>% ggplot +
  geom_density(aes(x = R), color = "red", fill = "red", alpha = 0.2) +
  geom_density(aes(x = R_prior), color = "blue", fill = "blue", alpha = 0.2) +
  theme_minimal()

w1p <- draws %>% ggplot +
  geom_density(aes(x = `w[1]`), color = "red", fill = "red", alpha = 0.2) +
  geom_density(aes(x = `w_prior[1]`), color = "blue", fill = "blue", alpha = 0.2) +
  theme_minimal()
w2p <- draws %>% ggplot +
  geom_density(aes(x = `w[2]`), color = "red", fill = "red", alpha = 0.2) +
  geom_density(aes(x = `w_prior[2]`), color = "blue", fill = "blue", alpha = 0.2) +
  theme_minimal()
w3p <- draws %>% ggplot +
  geom_density(aes(x = `w[3]`), color = "red", fill = "red", alpha = 0.2) +
  geom_density(aes(x = `w_prior[3]`), color = "blue", fill = "blue", alpha = 0.2) +
  theme_minimal()
w4p <- draws %>% ggplot +
  geom_density(aes(x = `w[4]`), color = "red", fill = "red", alpha = 0.2) +
  geom_density(aes(x = `w_prior[4]`), color = "blue", fill = "blue", alpha = 0.2) +
  theme_minimal()
w5p <- draws %>% ggplot +
  geom_density(aes(x = `w[5]`), color = "red", fill = "red", alpha = 0.2) +
  geom_density(aes(x = `w_prior[5]`), color = "blue", fill = "blue", alpha = 0.2) +
  theme_minimal()

w1p+w2p+w3p+w4p+w5p

draws %>% ggplot() +
  geom_point(aes(x = log_R, y = logit_c))

draws %>% ggplot() +
  geom_point(aes(x = log_R, y = `w[1]`))

draws %>% ggplot() +
  geom_point(aes(x = log_R, y = `w[2]`))
draws %>% ggplot() +
  geom_point(aes(x = log_R, y = `w[3]`))
draws %>% ggplot() +
  geom_point(aes(x = log_R, y = `w[4]`))
draws %>% ggplot() +
  geom_point(aes(x = log_R, y = `w[5]`))

```

Check chains
```{r}
tp1 <- draws %>% ggplot() +
  geom_line(aes(x = .iteration, y = logit_c, group = .chain, color = .chain)) +
  theme_minimal()
tp2 <- draws %>% ggplot() +
  geom_line(aes(x = .iteration, y = log_R, group = .chain, color = .chain)) +
  theme_minimal()
tp3 <- draws %>% ggplot() +
  geom_line(aes(x = .iteration, y = `w[1]`, group = .chain, color = .chain)) +
  theme_minimal()
tp4 <- draws %>% ggplot() +
  geom_line(aes(x = .iteration, y = `w[2]`, group = .chain, color = .chain)) +
  theme_minimal()
tp5 <- draws %>% ggplot() +
  geom_line(aes(x = .iteration, y = `w[3]`, group = .chain, color = .chain)) +
  theme_minimal()
tp6 <- draws %>% ggplot() +
  geom_line(aes(x = .iteration, y = `w[4]`, group = .chain, color = .chain)) +
  theme_minimal()
tp7 <- draws %>% ggplot() +
  geom_line(aes(x = .iteration, y = `w[5]`, group = .chain, color = .chain)) +
  theme_minimal()

(tp1 + tp2) / (tp3 + tp4) / (tp5 + tp6) / tp7

```

### Fit model to empirical data ###


```{r}
empirical_d <- read.csv("AlienData.txt")
```

Preprocess empirical data

```{r}
emp_d <- empirical_d %>% 
  filter(session == 1) %>% 
  mutate("f1" = substr(stimulus, 1, 1),
         "f2" = substr(stimulus, 2, 2),
         "f3" = substr(stimulus, 3, 3),
         "f4" = substr(stimulus, 4, 4),
         "f5" = substr(stimulus, 5, 5))
emp_d$subject <- paste(emp_d$condition, emp_d$subject, sep = "_")

```


Now we fit to each participant

```{r}
IDs <- unique(emp_d$subject)

draws_all <- data.frame()

for (ID in IDs){
  this_ID <- ID
  
  temp_d <- emp_d %>% filter(subject == this_ID)
  
  # prepare data
  w_prior_values = c(rep(0.2, 5))
  ntrials = length(temp_d$f1)
  nfeats = 5
  stimuli = array(NA, dim = c(ntrials, nfeats))
  stimuli[,1] = temp_d$f1
  stimuli[,2] = temp_d$f2
  stimuli[,3] = temp_d$f3
  stimuli[,4] = temp_d$f4
  stimuli[,5] = temp_d$f5
  
  temp_data <- list(
    ntrials = ntrials,
    nfeats = nfeats,
    stimuli = stimuli,
    response = temp_d$response,
    dangerous = temp_d$dangerous,
    nutricious = temp_d$nutricious,
    w_prior_values = w_prior_values
    )
  
  # fit model
  
  temp_samples <- gcm_kalman_mod$sample(
    data = temp_data,
    seed = 123,
    chains = 3,
    parallel_chains = 3,
    iter_warmup = 1000,
    iter_sampling = 3000,
    refresh = 0,
    max_treedepth = 10,
    adapt_delta = 0.80
  )
  
  temp_draws <- as_draws_df(temp_samples$draws()) # get draws df
  
  temp_draws$ID = this_ID
  
  temp_draws <- temp_draws %>% select(c, R, logit_c, log_R, c_prior, R_prior, logit_c_prior, log_R_prior,
                                      `w[1]`, `w[2]`, `w[3]`, `w[4]`, `w[5]`, .chain, .iteration, ID)
  
  draws_all <- rbind(draws_all, temp_draws) # save the draws from this participant
  
  print("Participant")
  print(ID)
  print("completed")
}

## Error: cannot allocate vector of size 431.7 Mb
## It got through 14 participants
```
save draws_all

```{r}
save(draws_all, file="draws_all_52subs.RData")
```



Plot posteriors

```{r}
library(ggridges)


draws_all$ID <- as.factor(draws_all$ID)
draws_all %>% ggplot() +
  geom_density_ridges(aes(x = R, y = ID, fill = ID), rel_min_height = 0.01) +
  xlim(c(0,5)) +
  theme_minimal()

draws_all %>% ggplot() +
  geom_density_ridges(aes(x = c, y = ID, fill = ID), rel_min_height = 0.01) +
  xlim(c(2.5, 4.2)) +
  theme_minimal()

w1p2 <- draws_all %>% ggplot() +
  geom_density_ridges(aes(x = `w[1]`, y = ID, fill = ID)) +
  xlim(c(0,1)) +
  theme_minimal()
w2p2 <- draws_all %>% ggplot() +
  geom_density_ridges(aes(x = `w[2]`, y = ID, fill = ID)) +
  xlim(c(0,1)) +
  theme_minimal()
w3p2 <- draws_all %>% ggplot() +
  geom_density_ridges(aes(x = `w[3]`, y = ID, fill = ID)) +
  xlim(c(0,1)) +
  theme_minimal()
w4p2 <- draws_all %>% ggplot() +
  geom_density_ridges(aes(x = `w[4]`, y = ID, fill = ID)) +
  xlim(c(0,1)) +
  theme_minimal()
w5p2 <- draws_all %>% ggplot() +
  geom_density_ridges(aes(x = `w[5]`, y = ID, fill = ID)) +
  xlim(c(0,1)) +
  theme_minimal()

w1p2
w2p2
w3p2
w4p2
w5p2



```






