---
title: "Assignment 3 code"
author: "Andreas"
date: "2024-04-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(gtools)
library(tidyverse)
library(cmdstanr)
library(posterior)
library(loo)
library(patchwork)
```


Assignment 3

Analyzing real world data from the social conformity experiment

In this markdown
- Data is simulated (for one participant)
- Simple Bayes and Weighted Bayes models are fit to the simulated data
- The models are fitted to the real data (cogsci students, pre-pandemic)
- Model quality checks
- Model comparison


### Simulate data ###

Define functions to generate the data
```{r}

simpleBayes_f <- function(ntrials, bias){
  
  FirstRating <- floor(runif(n = ntrials, min = 1, max = 9)) #first ratings sampled randomly
  GroupRating <- floor(runif(n = ntrials, min = 1, max = 9)) #group ratings sampled randomly
  
  p_FirstRating <- FirstRating / 9 # Changing ratings to a ]0,1[ space (probability)
  p_GroupRating <- GroupRating / 9
  
  l_FirstRating <- logit(p_FirstRating) # Changing ratings to log-odds space
  l_GroupRating <- logit(p_GroupRating)
  
  l_SecondRating <- bias + l_FirstRating + l_GroupRating # Compute second rating using Simple Bayes
  
  SecondRating <- round(inv.logit(l_SecondRating) * 9) # Turn back to original scale
  
  SecondRating <- ifelse(SecondRating < 0.5, 1,
                         ifelse(SecondRating > 8.5, 8, SecondRating)) # Bounded ratings in [1, 8]
  
  return(list(FirstRating=FirstRating, GroupRating=GroupRating, SecondRating=SecondRating))
  
}


weightedBayes_f <- function(ntrials, bias, w1, w2){
  
  w1 <- (w1 - 0.5)*2 # Changing waits from literature scale [0.5, 1] to suitable scale [0, 1]
  w2 <- (w2 - 0.5)*2
  
  FirstRating <- floor(runif(n = ntrials, min = 1, max = 9)) #first ratings sampled randomly
  GroupRating <- floor(runif(n = ntrials, min = 1, max = 9)) #group ratings sampled randomly
  
  p_FirstRating <- FirstRating / 9 # Changing ratings to a ]0,1[ space (probability)
  p_GroupRating <- GroupRating / 9
  
  l_FirstRating <- logit(p_FirstRating) # Changing ratings to log-odds space
  l_GroupRating <- logit(p_GroupRating)
  
  l_SecondRating <- bias + l_FirstRating * w1 + l_GroupRating * w2 # Compute second rating using Weighted Bayes
  
  SecondRating <- round(inv.logit(l_SecondRating) * 9) # Turn back to original scale
  
  SecondRating <- ifelse(SecondRating < 0.5, 1,
                         ifelse(SecondRating > 8.5, 8, SecondRating)) # Bounded ratings in [1, 8]
  
  return(list(FirstRating=FirstRating, GroupRating=GroupRating, SecondRating=SecondRating))
  
}

```


Environment
```{r}
ntrials = 80
bias = 1
w1 = 0.7
w2 = 0.9
```


Inspect the Simple Bayes simulation

```{r}
data1 <- simpleBayes_f(ntrials, bias)

df1 <- tibble(FirstRating = data1$FirstRating, GroupRating = data1$GroupRating, SecondRating = data1$SecondRating, trial = seq(ntrials))


df1 %>% ggplot() +
  geom_point(aes(trial, FirstRating), color = "Blue") +
  geom_point(aes(trial, GroupRating), color = "Red") +
  geom_point(aes(trial, SecondRating), color = "Green") +
  ggtitle("Simple Bayes sim. Blue = First, Red = Group, Green = Second")

p1 <- df1 %>% ggplot() +
  geom_point(aes(FirstRating, GroupRating, color = SecondRating)) +
  ggtitle("Simple Bayes sim. Second rating as function of first and group ratings")

p1 + scale_color_gradient(low = "blue", high = "red")

```

Inspect the weighted Bayes simulation

```{r}
data2 <- weightedBayes_f(ntrials, bias, w1, w2)

df2 <- tibble(FirstRating = data2$FirstRating, GroupRating = data2$GroupRating, SecondRating = data2$SecondRating, trial = seq(ntrials))


df2 %>% ggplot() +
  geom_point(aes(trial, FirstRating), color = "Blue") +
  geom_point(aes(trial, GroupRating), color = "Red") +
  geom_point(aes(trial, SecondRating), color = "Green") +
  ggtitle("Simple Bayes sim. Blue = First, Red = Group, Green = Second")


p2 <- df2 %>% ggplot() +
  geom_point(aes(FirstRating, GroupRating, color = SecondRating)) +
  ggtitle("Weighted Bayes sim. Second rating as function of first and group ratings")

p2 + scale_color_gradient(low = "blue", high = "red")
```



### Fit models to simulated data ###


Simple Bayes model
```{r}
file1 = file.path("C:/Users/andre/OneDrive - Aarhus universitet/Cognitive Science/Kandidat/Advanced Cognitive Modeling/Assignments/Ass3/ass_3_simpleBayes.stan")

mod1 = cmdstan_model(file1)

data <- list(n = ntrials,
             FirstRating = df1$FirstRating,
             GroupRating = df1$GroupRating,
             SecondRating = df1$SecondRating)

samples1 <- mod1$sample(
  data = data,
  seed = 123,
  chains = 3,
  iter_warmup = 1000,
  iter_sampling = 3000,
  refresh = 0,
  max_treedepth = 10,
  adapt_delta = 0.80
)

draws1 <- as_draws_df(samples1$draws())

# Plot Prior Posterior update check for bias
draws1 %>%
  ggplot() +
  geom_density(aes(x=bias_prior), color = "blue", fill = "blue", alpha = 0.2) +
  geom_density(aes(x=bias), color = "red", fill = "red", alpha = 0.2) +
  geom_vline(xintercept = bias) +
  xlab("Bias") +
  ylab("Density") +
  ggtitle("Prior (blue) Posterior (red) Update Check - Bias") +
  theme_minimal()

# Plot Prior Posterior update check for error
draws1 %>%
  ggplot() +
  geom_density(aes(x=error_prior), color = "blue", fill = "blue", alpha = 0.2) +
  geom_density(aes(x=error), color = "red", fill = "red", alpha = 0.2) +
  geom_vline(xintercept = 0) +
  xlab("Error") +
  ylab("Density") +
  ggtitle("Prior (blue) Posterior (red) Update Check - Error") +
  theme_minimal()



# Check chains

draws1 %>% mutate(.chain = as.factor(.chain)) %>%  ggplot() +
  geom_line(aes(.iteration, bias, color = .chain)) +
  theme_minimal()

draws1 %>% mutate(.chain = as.factor(.chain)) %>%  ggplot() +
  geom_line(aes(.iteration, error, color = .chain)) +
  theme_minimal()


```


Weighted Bayes model (V1)

Don't run the below chunk, Stan model is not entirely correct

```{r}
# file2 = file.path("C:/Users/andre/OneDrive - Aarhus universitet/Cognitive Science/Kandidat/Advanced Cognitive Modeling/Assignments/Ass3/ass_3_weightedBayes.stan")
# 
# mod2 = cmdstan_model(file2)
# 
# data <- list(n = ntrials,
#              FirstRating = df2$FirstRating,
#              GroupRating = df2$GroupRating,
#              SecondRating = df2$SecondRating)
# 
# samples2 <- mod2$sample(
#   data = data,
#   seed = 123,
#   chains = 3,
#   iter_warmup = 1000,
#   iter_sampling = 3000,
#   refresh = 0,
#   max_treedepth = 10,
#   adapt_delta = 0.80
# )
# 
# draws2 <- as_draws_df(samples2$draws())
# 
# # Plot Prior Posterior update check for w1
# draws2 %>%
#   ggplot() +
#   geom_density(aes(x=w1_prior), color = "blue", fill = "blue", alpha = 0.2) +
#   geom_density(aes(x=w1), color = "red", fill = "red", alpha = 0.2) +
#   geom_vline(xintercept = w1) +
#   xlab("w1") +
#   ylab("Density") +
#   ggtitle("Prior (blue) Posterior (red) Update Check - w1") +
#   theme_minimal()
# 
# # Plot Prior Posterior update check for logit_w1 (the scale it was sampled on)
# draws2 %>%
#   ggplot() +
#   geom_density(aes(x=logit_w1_prior), color = "blue", fill = "blue", alpha = 0.2) +
#   geom_density(aes(x=logit_w1), color = "red", fill = "red", alpha = 0.2) +
#   #geom_vline(xintercept = w1) +
#   xlab("logit w1") +
#   ylab("Density") +
#   ggtitle("Prior (blue) Posterior (red) Update Check - logit w1") +
#   theme_minimal()
# 
# # Plot Prior Posterior update check for w2
# draws2 %>%
#   ggplot() +
#   geom_density(aes(x=w2_prior), color = "blue", fill = "blue", alpha = 0.2) +
#   geom_density(aes(x=w2), color = "red", fill = "red", alpha = 0.2) +
#   geom_vline(xintercept = w2) +
#   xlab("w2") +
#   ylab("Density") +
#   ggtitle("Prior (blue) Posterior (red) Update Check - w2") +
#   theme_minimal()
# 
# # Plot Prior Posterior update check for logit_w2 (the scale it was sampled on)
# draws2 %>%
#   ggplot() +
#   geom_density(aes(x=logit_w2_prior), color = "blue", fill = "blue", alpha = 0.2) +
#   geom_density(aes(x=logit_w2), color = "red", fill = "red", alpha = 0.2) +
#   #geom_vline(xintercept = w1) +
#   xlab("logit w2") +
#   ylab("Density") +
#   ggtitle("Prior (blue) Posterior (red) Update Check - logit w2") +
#   theme_minimal()
# 
# 
# # Plot Prior Posterior update check for bias
# draws2 %>%
#   ggplot() +
#   geom_density(aes(x=bias_prior), color = "blue", fill = "blue", alpha = 0.2) +
#   geom_density(aes(x=bias), color = "red", fill = "red", alpha = 0.2) +
#   geom_vline(xintercept = bias) +
#   xlab("Bias") +
#   ylab("Density") +
#   ggtitle("Prior (blue) Posterior (red) Update Check - Bias") +
#   theme_minimal()
# 
# # Plot Prior Posterior update check for error
# draws2 %>%
#   ggplot() +
#   geom_density(aes(x=error_prior), color = "blue", fill = "blue", alpha = 0.2) +
#   geom_density(aes(x=error), color = "red", fill = "red", alpha = 0.2) +
#   geom_vline(xintercept = 0) +
#   xlab("Error") +
#   ylab("Density") +
#   ggtitle("Prior (blue) Posterior (red) Update Check - Error") +
#   theme_minimal()
# 
# 
# 
# # Check chains
# 
# draws2 %>% mutate(.chain = as.factor(.chain)) %>%  ggplot() +
#   geom_line(aes(.iteration, bias, color = .chain)) +
#   theme_minimal()
# 
# draws2 %>% mutate(.chain = as.factor(.chain)) %>%  ggplot() +
#   geom_line(aes(.iteration, error, color = .chain)) +
#   theme_minimal()
# 
# draws2 %>% mutate(.chain = as.factor(.chain)) %>%  ggplot() +
#   geom_line(aes(.iteration, w1, color = .chain)) +
#   theme_minimal()
# 
# draws2 %>% mutate(.chain = as.factor(.chain)) %>%  ggplot() +
#   geom_line(aes(.iteration, w2, color = .chain)) +
#   theme_minimal()


```



Weighted Bayes Model (V2 - Same output as V1, but parameters specified and transformed in reverse order)

```{r}
file3 = file.path("C:/Users/andre/OneDrive - Aarhus universitet/Cognitive Science/Kandidat/Advanced Cognitive Modeling/Assignments/Ass3/ass_3_weightedBayes_V2.stan")

mod3 = cmdstan_model(file3)

data <- list(n = ntrials,
             FirstRating = df2$FirstRating,
             GroupRating = df2$GroupRating,
             SecondRating = df2$SecondRating)

samples3 <- mod3$sample(
  data = data,
  seed = 123,
  chains = 3,
  iter_warmup = 1000,
  iter_sampling = 3000,
  refresh = 0,
  max_treedepth = 10,
  adapt_delta = 0.80
)

draws3 <- as_draws_df(samples3$draws())

# Plot Prior Posterior update check for w1
draws3 %>%
  ggplot() +
  geom_density(aes(x=w1_prior), color = "blue", fill = "blue", alpha = 0.2) +
  geom_density(aes(x=w1_posterior), color = "red", fill = "red", alpha = 0.2) +
  geom_vline(xintercept = w1) +
  xlab("w1") +
  ylab("Density") +
  ggtitle("Prior (blue) Posterior (red) Update Check - w1") +
  theme_minimal()

# Plot Prior Posterior update check for logit_w1 (the scale it was sampled on)
draws3 %>%
  ggplot() +
  geom_density(aes(x=logit_w1_prior), color = "blue", fill = "blue", alpha = 0.2) +
  geom_density(aes(x=logit_w1), color = "red", fill = "red", alpha = 0.2) +
  #geom_vline(xintercept = w1) +
  xlab("logit w1") +
  ylab("Density") +
  ggtitle("Prior (blue) Posterior (red) Update Check - logit w1") +
  theme_minimal()

# Plot Prior Posterior update check for w2
draws3 %>%
  ggplot() +
  geom_density(aes(x=w2_prior), color = "blue", fill = "blue", alpha = 0.2) +
  geom_density(aes(x=w2_posterior), color = "red", fill = "red", alpha = 0.2) +
  geom_vline(xintercept = w2) +
  xlab("w2") +
  ylab("Density") +
  ggtitle("Prior (blue) Posterior (red) Update Check - w2") +
  theme_minimal()

# Plot Prior Posterior update check for logit_w2 (the scale it was sampled on)
draws3 %>%
  ggplot() +
  geom_density(aes(x=logit_w2_prior), color = "blue", fill = "blue", alpha = 0.2) +
  geom_density(aes(x=logit_w2), color = "red", fill = "red", alpha = 0.2) +
  #geom_vline(xintercept = w1) +
  xlab("logit w2") +
  ylab("Density") +
  ggtitle("Prior (blue) Posterior (red) Update Check - logit w2") +
  theme_minimal()


# Plot Prior Posterior update check for bias
draws3 %>%
  ggplot() +
  geom_density(aes(x=bias_prior), color = "blue", fill = "blue", alpha = 0.2) +
  geom_density(aes(x=bias), color = "red", fill = "red", alpha = 0.2) +
  geom_vline(xintercept = bias) +
  xlab("Bias") +
  ylab("Density") +
  ggtitle("Prior (blue) Posterior (red) Update Check - Bias") +
  theme_minimal()

# Plot Prior Posterior update check for error
draws3 %>%
  ggplot() +
  geom_density(aes(x=error_prior), color = "blue", fill = "blue", alpha = 0.2) +
  geom_density(aes(x=error), color = "red", fill = "red", alpha = 0.2) +
  geom_vline(xintercept = 0) +
  xlab("Error") +
  ylab("Density") +
  ggtitle("Prior (blue) Posterior (red) Update Check - Error") +
  theme_minimal()



# Check chains

tp1 <- draws3 %>% mutate(.chain = as.factor(.chain)) %>%  ggplot() +
  geom_line(aes(.iteration, bias, color = .chain)) +
  theme_minimal()

tp2 <- draws3 %>% mutate(.chain = as.factor(.chain)) %>%  ggplot() +
  geom_line(aes(.iteration, error, color = .chain)) +
  theme_minimal()

tp3 <- draws3 %>% mutate(.chain = as.factor(.chain)) %>%  ggplot() +
  geom_line(aes(.iteration, logit_w1, color = .chain)) +
  theme_minimal()

tp4 <- draws3 %>% mutate(.chain = as.factor(.chain)) %>%  ggplot() +
  geom_line(aes(.iteration, logit_w2, color = .chain)) +
  theme_minimal()

(tp1 + tp2) / (tp3 + tp4)


```

### Fit to real data ###


Load the data
```{r}
cogsci_d <- read_csv("cogsci_clean.csv")
```

Clean the data
```{r}
# Detect ratings below 1 or above 8
length(cogsci_d$FirstRating[cogsci_d$FirstRating < 1])
length(cogsci_d$FirstRating[cogsci_d$FirstRating > 8])

length(cogsci_d$SecondRating[cogsci_d$SecondRating < 1])
length(cogsci_d$SecondRating[cogsci_d$SecondRating > 8])

length(cogsci_d$GroupRating[cogsci_d$GroupRating < 1])
length(cogsci_d$GroupRating[cogsci_d$GroupRating > 8])


# There are 0's in Group ratings. What to do with these?

cogsci_d <- cogsci_d %>% filter(GroupRating > 0)

```

Inspect distribution of second ratings pr participant
```{r}
cogsci_d$ID <- as.factor(cogsci_d$ID)
cogsci_d %>% ggplot() +
  geom_density(aes(SecondRating, color = ID, fill = ID), alpha = 0.1) +
  theme_minimal()

# One participant rated 1 most of the times, one almost always rated 5
```


Fit Simple Bayes Model to the real data

```{r}
IDs <- unique(cogsci_d$ID)

draws_all <- data.frame()

for (ID in IDs){
  this_ID <- ID # Why the f is this necessary???
  single_d <- cogsci_d %>% filter(ID == this_ID)
  
  temp_data <- list(n = length(single_d$FirstRating),
                    FirstRating = single_d$FirstRating,
                    GroupRating = single_d$GroupRating,
                    SecondRating = single_d$SecondRating)

  temp_samples <- mod1$sample(
    data = temp_data,
    seed = 123,
    chains = 3,
    iter_warmup = 1000,
    iter_sampling = 3000,
    refresh = 0,
    max_treedepth = 10,
    adapt_delta = 0.80
  )
  
  temp_draws <- as_draws_df(temp_samples$draws())
  
  temp_draws$ID <- ID
  
  draws_all <- rbind(draws_all, temp_draws)
}

```

Plot all the posteriors for Bias and Error
```{r}
draws_all$ID <- as.factor(draws_all$ID)

draws_all %>%
  ggplot() +
  geom_density(aes(x=bias, color = ID, fill = ID), alpha = 0.2) +
  xlab("Bias") +
  ylab("Density") +
  ggtitle("Bias Posteriors") +
  theme_minimal()

draws_all %>%
  ggplot() +
  geom_density(aes(x=error, color = ID, fill = ID), alpha = 0.2) +
  xlab("Error") +
  ylab("Density") +
  ggtitle("Error Posteriors") +
  theme_minimal()
```


Fit Weighted Bayes Model to real data

```{r}
draws_all2 <- data.frame()

for (ID in IDs){
  this_ID <- ID # Why the f is this necessary???
  single_d <- cogsci_d %>% filter(ID == this_ID)
  
  temp_data <- list(n = length(single_d$FirstRating),
                    FirstRating = single_d$FirstRating,
                    GroupRating = single_d$GroupRating,
                    SecondRating = single_d$SecondRating)

  temp_samples <- mod3$sample(
    data = temp_data,
    seed = 123,
    chains = 3,
    iter_warmup = 1000,
    iter_sampling = 3000,
    refresh = 0,
    max_treedepth = 10,
    adapt_delta = 0.80
  )
  
  temp_draws <- as_draws_df(temp_samples$draws())
  
  temp_draws$ID <- ID
  
  draws_all2 <- rbind(draws_all2, temp_draws)
}
```


Plot all the posteriors for Bias, Error, w1, and w2

```{r}
draws_all2$ID <- as.factor(draws_all2$ID)

draws_all2 %>%
  ggplot() +
  geom_density(aes(x=bias, color = ID, fill = ID), alpha = 0.2) +
  xlab("Bias") +
  ylab("Density") +
  ggtitle("Bias Posteriors") +
  theme_minimal()

draws_all2 %>% filter(ID != 27) %>% 
  ggplot() +
  geom_density(aes(x=error, color = ID, fill = ID), alpha = 0.2) +
  xlab("Error") +
  ylab("Density") +
  ggtitle("Error Posteriors") +
  theme_minimal()

draws_all2 %>%
  ggplot() +
  geom_density(aes(x=logit_w1, fill = ID), color = "blue", alpha = 0.1) +
  geom_density(aes(x=logit_w2, fill = ID), color = "red", alpha = 0.1) +
  xlab("Logit Weight") +
  ylab("Density") +
  ggtitle("Logit Weight Posteriors") +
  theme_minimal()


# We plot on the [0.5, 1] scale but remove participant 27, as their posterior for weight is almost exactly 0.5

draws_all2 %>% filter(ID != 27) %>% 
  ggplot() +
  geom_density(aes(x=w1_posterior, fill = ID), color = "blue", alpha = 0.1) +
  geom_density(aes(x=w2_posterior, fill = ID), color = "red", alpha = 0.1) +
  xlab("Weight") +
  ylab("Density") +
  ggtitle("Weight Posteriors") +
  theme_minimal()

```


### Model Comparison ###

In order to compare the two models, we need a single fit for each model. Hence, we fit to the entire dataset.


Remove participant 27 to not mess with the weight estimates
```{r}
cogsci_d <- cogsci_d %>% filter(ID != 27)
```


Fit the models to the data
```{r}
data <- list(n = length(cogsci_d$SecondRating),
             FirstRating = cogsci_d$FirstRating,
             GroupRating = cogsci_d$GroupRating,
             SecondRating = cogsci_d$SecondRating)

simpleBayes_fit <- mod1$sample(
  data = data,
  seed = 123,
  chains = 3,
  iter_warmup = 1000,
  iter_sampling = 3000,
  refresh = 0,
  max_treedepth = 10,
  adapt_delta = 0.80
)

weightedBayes_fit <- mod3$sample(
  data = data,
  seed = 123,
  chains = 3,
  iter_warmup = 1000,
  iter_sampling = 3000,
  refresh = 0,
  max_treedepth = 10,
  adapt_delta = 0.80
)

```


Create Loo objects
```{r}
Loo_simpleBayes <- simpleBayes_fit$loo(save_psis = TRUE, cores = 4)

Loo_weightedBayes <- weightedBayes_fit$loo(save_psis = TRUE, cores = 4)
```

PSIS diagnostic plot
```{r}
plot(Loo_simpleBayes)
plot(Loo_weightedBayes)
```



Create differentiated elpd
```{r}
elpd <- tibble(
  n = seq(length(cogsci_d$SecondRating)),
  differentiated_elpd = 
  Loo_simpleBayes$pointwise[, "elpd_loo"] - 
  Loo_weightedBayes$pointwise[, "elpd_loo"]
  )
```


Plot the differentiated elpd
```{r}
elpd %>% ggplot(aes(x = n, y = differentiated_elpd)) +
  geom_point(alpha = .1) +
  #xlim(.5,1.01) +
  #ylim(-1.5,1.5) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  theme_minimal()
```



Make Loo comparison
```{r}
loo_compare(Loo_simpleBayes, Loo_weightedBayes)
```

Get model weights (how much do we gain from using a second model to explain the data - ish)
```{r}
loo_model_weights(list(Loo_weightedBayes, Loo_simpleBayes))
```

